Quantitative Benchmarking for Automatically Generated Conjectures

### 2018-05-30 TEBenchmarkPaper/article.tex
Fix some TEB issues, e.g. explaining the different averages

\begin{itemize}
\item \emph{Precision} is the proportion of a tool's output which appears in
  the ground truth. This penalises overly-liberal tools which output a large
  number of properties in the hope that some turn out to be ``good''.
\item \emph{Recall} is the proportion of the ground truth which appears in the
  tool's output. This penalises overly-conservative tools which generate very
  few properties as a way to avoid ``bad'' ones.
\end{itemize}

@@

QuickSpec emerged from work on the QuickCheck software testing framework for the
Haskell programming language~\cite{claessen2011quickcheck}. As well as
conjecturing properties of Haskell definitions for HipSpec, it has been applied
to Isabelle definitions via the Hipster tool~\cite{Hipster}. Like IsaCoSy,
QuickSpec takes a signature and enumerates well-typed terms, collecting together
those of the same type into equivalence classes, assuming them to be
equal. QuickCheck is then used to find counterexamples to this assumption, which
it does by randomly instantiating the variables and comparing the resulting
closed expressions. Any equivalence class whose elements don't compare equal are
split up, and the process is repeated with new random values.

@@

\begin{itemize}
\item For each problem, there should be a clear distinction between the
  statement to be proved and the definitions involved, such that the two can be
  easily and meaningfully separated. This rules out problem sets like those of
  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
  functions, whose behaviour is \emph{implicit} in the logical structure of the
  theorem statement but not separately \emph{defined}.
\item Definitions should be translatable into a form suitable for the MTE tools
  under study. Our application in $\S$\ref{sec:application} requires Haskell and
  Isabelle translations, and also benefits from having definitions be strongly
  typed.
\item The problem set should be relevant to the desired domain. Our focus on
  functional programming requires higher-order functions and inductively defined
  datatypes, which rules out first-order languages/logics (such as
  TPTP~\cite{sutcliffe2009tptp}).
\item The problem set should ideally contain \emph{every} ``interesting''
  property involving its included definitions, since non-membership in the
  ground truth will be treated as being ``uninteresting''. More realistically,
  we should aim for each definition to have \emph{multiple} properties; rather
  than a mixture of unrelated problems, cherry-picked from different domains.
\item Larger problem set are preferable as they give more robust statistics, all
  else being equal (i.e. when this does not sacrifice quality).
\end{itemize}

@@

It is important to ensure that there are no duplicate definitions: we are only
concerned with the \emph{logical} content of the input, not the more arbitrary
aspects like the names of functions. For example, consider a problem set which
includes a statement of commutativity for a \texttt{plus} function, and of
associativity for an \texttt{add} function, where the definitions of
\texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would expect an MTE
tool to either conjecture commutativity and associativity for \emph{both}
functions, or for \emph{neither} function, since they are logically equivalent.
Yet a na\"ive precision/recall analysis would treat commutativity of
\texttt{add} and associativity of \texttt{plus} as \emph{uninteresting}, since
they don't appear in the ground truth.

@@

To solve both of these problems we instead \emph{sample} a subset of
definitions.\footnote{This could be done randomly, but for reproducibility we
  use a deterministic order (based on cryptographic hashes, to discourage
  ``cheating'').} Given a sample size, we choose a subset of that many
definitions, and provide only those as input to the tool. We generate a
corresponding ground truth by selecting those properties from the corpus which
``depend on'' (contain references to) \emph{only} the definitions in that
sample. Transitive dependencies aren't required (e.g. a property involving only
a \texttt{times} function would not depend on a \texttt{plus} function, even if
\texttt{plus} occurs in the definition of \texttt{times}).

@@

Instead, we only allow a sample if it contains all dependencies of at least one
property. We could do this using rejection sampling, but it is more efficient to
pick a property, weighted in proportion to their number of dependencies
(ignoring those with more dependencies than our sample size). That property's
dependencies become our sample, padded up to the required size with uniform
choices from the remaining definitions.

The ground truth for such samples is guaranteed to contain at least one property
(the one we picked), and hence the precision and recall will depend meaningfully
on the tool's output.

@@

We can summarise runtimes by choosing their median, as this is more robust than
the mean against long-running outliers, and hence represents performance for a
``typical'' input of that size. Unlike previous one-off evaluations, we can also
compute the \emph{spread} of our data, for example with the inter-quartile
range.

The precision and recall for multiple samples can be combined in two ways. We
can calculate their mean value, known as the \emph{average of the ratios}, to
find the expected precision and recall for runs of this size. Alternatively, we
can sum their numerators and denominators to get the \emph{ratio of the
  averages}: this normalises the output and ground truth cardinalities, allowing
precision and recall to be modelled as simple Bernoulli processes.

We prefer the latter, where each generated conjecture (wanted or not) is like a
coin toss whose bias is the precision; likewise, the ground truth properties
(found or not) are like coin tosses biased by the recall. Note that this simple
model does not explain how many conjectures the tool decides to output for each
run.

@@

To measure progress and encourage competition in the field, it is important to
compare the relative performance of different tools on the same task. Since the
aggregate statistics in the above summaries have obscured the details of
specific runs, any comparison based on them (such as comparing average recall)
would have very low statistical power. More direct comparisons can be made using
\emph{paired} tests, since for each individual sample we have measurements for
\emph{both} tools.

@@

For simplicity, we suggest comparing quality by pooling together results from
all samples where both tools finished successfully. Recall can be compared using
the standard McNemar test for paired samples~\cite{mcnemar1947note}: each
ground-truth property is categorised as either being found only by the first
tool, only by the second, by both or by neither. Those found by both or neither
are ignored, and the tool-specific numbers are tested for marginal homogeneity
(i.e. whether their differences give rise to similar totals).

@@

We chose the Tons of Inductive Problems (TIP) benchmark for our ground truth
since it satisfies the criteria of $\S$\ref{section:prep}: each benchmark
problem has standalone type and function definitions, making their separation
trivial; known examples from the software verification and inductive theorem
proving literature are included, ensuring relevance to those fields; the format
includes the higher-order functions and inductive datatypes we are interested
in; it is large enough to pose a challenge to current MTE tools; plus it is
accompanied by tooling to convert its custom format (an extension of
SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of languages, including Haskell and
Isabelle.

@@

TIP comes with software to translate its definitions into Haskell and Isabelle
code, including comparison functions and random data generators suitable for
QuickCheck. We translate all 269 unique definitions into a single module/theory
which is imported on each run of the tools, although only those functions which
appear in the current sample are included in the signature and explored.

@@

We benchmarked QuickSpec version 1 (0.9.6), a tool written in Haskell for
conjecturing equations involving Haskell functions, described in more detail in
$\S$\ref{sec:existing-tools}. In order to thoroughly benchmark QuickSpec, we
need to automate some of the decisions which are normally left up to the user:

@@

We compare the recall proportions by applying McNemar's test to only those
samples where both tools succeeded. We pooled these together from all sample
sizes to produce table~\ref{table:recall}, and found that the recall of
QuickSpec ($37\%$) is significantly higher than IsaCoSy ($25\%$) with a p value
of 0.0026.

\begin{table}
  \centering
  \begin{tabular}{ |r|l|l| }
    \hline
    & \bfseries Found by IsaCoSy & \bfseries Not found by IsaCoSy \\
    \hline
    \bfseries Found by QuickSpec     & 28 & 19 \\
    \hline
    \bfseries Not found by QuickSpec & 4  & 75 \\
    \hline
  \end{tabular}
  \caption{Contingency table for ground truth properties, combined from those
    samples where both QuickSpec and IsaCoSy finished successfully. The combined
    recall for QuickSpec is $37\%$ and for IsaCoSy is $25\%$.}
  \label{table:recall}
\end{table}

@@

\begin{table}
  \centering
  \begin{tabular}{ |r|l|l| }
    \hline
              & \bfseries Interesting & \bfseries Uninteresting \\
    \hline
    \bfseries IsaCoSy   & 32          & 137      \\
    \hline
    \bfseries QuickSpec & 47          & 301      \\
    \hline
  \end{tabular}
  \caption{Interesting and uninteresting conjectures generated by QuickSpec and
    IsaCoSy, combined from those samples where both both finished successfully.
    The combined precision for QuickSpec is $14\%$ and for IsaCoSy is $19\%$.}
  \label{table:precision}
\end{table}

@@

One notable weakness is that our methodology only makes sense in domains where
existing human knowledge surpasses that discovered by the machine. Any
\emph{truly novel} insights discovered during testing will not, by definition,
appear in any existing corpus, and we would in fact \emph{penalise} tools for
such output. We do not believe this to be a realistic problem for the time
being, although it does emphasise the continued importance of testing these
tools with real human users, rather than relying solely on artificial benchmarks
such as ours.

>>>>

\begin{itemize}
\item \emph{Precision} is the proportion of a tool's output which appears in
  the ground truth (the ratio of true positives to all positives). This
  penalises overly-liberal tools which output a large number of properties in
  the hope that some turn out to be ``good''.
\item \emph{Recall} is the proportion of the ground truth which appears in the
  tool's output (the ratio of true positives to actual positives). This
  penalises overly-conservative tools which generate very few properties as a
  way to avoid ``bad'' ones.
\end{itemize}

@@

QuickSpec emerged from work on the QuickCheck software testing framework for the
Haskell programming language~\cite{claessen2011quickcheck}. As well as
conjecturing properties of Haskell definitions for HipSpec, it has been applied
to Isabelle definitions via the Hipster tool~\cite{Hipster}. Like IsaCoSy,
QuickSpec takes a signature and enumerates well-typed terms. It collects
together those of the same type into equivalence classes, assuming them to be
equal, then uses QuickCheck to find counterexamples to this assumption by
randomly instantiating the variables and comparing the resulting closed
expressions. Any equivalence class whose elements don't compare equal are split
up, and the process is repeated with new random values.

@@

\begin{itemize}
\item For each problem, there should be a clear distinction between the
  statement to be proved and the definitions involved, such that the two can be
  easily and meaningfully separated. This rules out problem sets like those of
  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
  functions, whose behaviour is \emph{implicit} in the logical structure of the
  theorem statement but not separately \emph{defined}.
\item Definitions should be translatable into a form suitable for the MTE tools
  under study. Our application in $\S$\ref{sec:application} requires Haskell and
  Isabelle translations, and also benefits from having definitions be strongly
  typed.
\item The problem set should be relevant to the desired domain. We focus on pure
  functional programming, which requires higher-order functions and inductively
  defined datatypes, which rules out first-order languages/logics (such as
  TPTP~\cite{sutcliffe2009tptp}).
\item The problem set should ideally contain \emph{every} ``interesting''
  property involving its included definitions, since non-membership in the
  ground truth will be treated as being ``uninteresting''. More realistically,
  we should aim for each definition to have \emph{multiple} properties; rather
  than a mixture of unrelated problems, cherry-picked from different domains.
\item Larger problem sets are preferable as they give more robust statistics, all
  else being equal (i.e. when this does not sacrifice quality).
\end{itemize}

@@

It is important to ensure that there are no duplicate definitions: we are only
concerned with the \emph{logical} content of the input, not the more arbitrary
aspects of their presentation like the names of functions. For example, consider
a problem set which includes a statement of commutativity for a \texttt{plus}
function, and of associativity for an \texttt{add} function, where the
definitions of \texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would
expect an MTE tool to either conjecture commutativity and associativity for
\emph{both} functions, or for \emph{neither} function, since they are logically
equivalent. Yet a na\"ive precision/recall analysis would treat commutativity of
\texttt{add} and associativity of \texttt{plus} as \emph{uninteresting}, since
they don't appear in the ground truth.

@@

To solve both of these problems we instead \emph{sample} a subset of
definitions.\footnote{This could be done randomly, but for reproducibility we
  use a deterministic order based on cryptographic hashes.} Given a sample size,
we choose a subset of that many definitions, and provide only those as input to
the tool. We generate a corresponding ground truth by selecting those properties
from the corpus which ``depend on'' (contain references to) \emph{only} the
definitions in that sample. Transitive dependencies aren't required (e.g. a
property involving only a \texttt{times} function would not depend on a
\texttt{plus} function, even if \texttt{plus} occurs in the definition of
\texttt{times}).

@@

Other measurements could be used for such cases, but our approach is to avoid
them by only allowing a sample if it contains all dependencies of at least one
property. We could do this using rejection sampling, but it is more efficient to
pick a property, weighted in proportion to their number of dependencies
(ignoring those with more dependencies than our sample size). That property's
dependencies become our sample, padded up to the required size with uniform
choices from the remaining definitions.

The ground truth for such samples is guaranteed to contain at least one property
(the one we picked), and hence the precision and recall will depend meaningfully
on the tool's output. One downside of this restriction is that we limit the
number of possible samples to measure; this is significant for sample sizes less
than 3.

@@

We can summarise runtimes by choosing their median, as this is more robust than
the mean against long-running outliers, and hence represents performance for a
``typical'' input of that size. Since our methodology measures performance over
many samples, we can also compute the \emph{spread} of our data, for example the
inter-quartile range. This is not possible using the one-off measurements common
in prior evaluation methods.

There are two ways to average the measured precision and (separately) recall for
samples of a particular size. Since these quantities are ratios, their mean
values are known as the \emph{average of the ratios} (AOR), which we can define
for $S$ samples indexed by $1 \leq i \leq S$ as follows:

\begin{align*}
       \overline{\text{precision}}_{AOR}
    &= \frac{1}{S} \sum_i{\text{precision}_i} \\
    &= \frac{1}{S} \sum_i{
         \frac{\#\text{interesting}_i}
              {\#\text{generated}_i}}         \\[10pt]
       \overline{\text{recall}}_{AOR}
    &= \frac{1}{S} \sum_i{\text{recall}_i} \\
    &= \frac{1}{S} \sum_i{
         \frac{\#\text{interesting}_i}
              {\#\text{groundtruth}_i}}
\end{align*}

We interpret these as the \emph{expected} values of precision and recall for
samples of this size, i.e. the expected quality of the \emph{set} of conjectures
that will be generated from such a sample.

The alternative is to calculate \emph{pooled} averages, by combining the
(multi)sets from each sample before taking the ratios. If we multiply the
numerator and denominator by a factor $\frac{1}{S}$ we find that this is
equivalent to using the mean set sizes, hence this is known as the
\emph{ratio of the averages} (ROA):

\begin{align*}
       \overline{\text{precision}}_{ROA}
    &= \frac{\sum_i{\#\text{interesting}_i}}
            {\sum_i{\#\text{generated}_i}}                              \\
    &= \frac{\frac{1}{S} \sum_i{\#\text{interesting}_i}}
            {\frac{1}{S} \sum_i{\#\text{generated}_i}}                  \\
    &= \frac{\hspace{5pt} \overline{\#\text{interesting}} \hspace{5pt}}
            {\hspace{5pt} \overline{\#\text{generated}}   \hspace{5pt}} \\[10pt]
       \overline{\text{recall}}_{ROA}
    &= \frac{\sum_i{\#\text{interesting}_i}}
            {\sum_i{\#\text{groundtruth}_i}}                            \\
    &= \frac{\frac{1}{S} \sum_i{\#\text{interesting}_i}}
            {\frac{1}{S} \sum_i{\#\text{groundtruth}_i}}                \\
    &= \frac{\hspace{5pt} \overline{\#\text{interesting}} \hspace{5pt}}
            {\hspace{5pt} \overline{\#\text{groundtruth}} \hspace{5pt}}
\end{align*}

Larger sets contribute more to a ratio of averages, unlike the mean which treats
each set equally. This weighting gives us expected values of \emph{conjectures}
rather than sets: the ROA for precision is the expected chance of a generated
conjecture being interesting; the ROA for recall is the expected chance that a
ground truth property will be generated. Note that both of these averaging
methods deal with \emph{fractions}: we may also want to know the \emph{absolute}
number of conjectures to expect, which is the mean output size
$\overline{\#\text{generated}}$.

If we instead want to combine precision values \emph{with} recall values (either
individually or pooled) we can calculate their \emph{F-score}, defined as:

\begin{equation*}
  F = 2 \times \frac{\text{precision} \times \text{recall}}
                    {\text{precision} +      \text{recall}}
\end{equation*}

This is the harmonic mean of precision and recall, ranging between 0 and 1. The
F-score summarises performance into a single number, which is useful for
purposes like ranking, but it obscures insights we might gain from looking at
precision and recall separately (e.g. whether a tool is generating too much
output or too little).

@@

To measure progress and encourage competition in the field, it is important to
compare the relative performance of different tools on the same task. Since the
aggregate statistics in the above summaries have obscured the details of
specific runs, any comparison based on them (such as comparing mean recall)
would have very low statistical power. More direct comparisons can be made using
\emph{paired} tests, since for each individual sample we have measurements for
\emph{both} tools.

@@

For simplicity, we suggest comparing quality by pooling results from all samples
where both tools finished successfully. Recall can be compared using the
standard McNemar test for paired samples~\cite{mcnemar1947note}: each
ground-truth property is categorised as either being found only by the first
tool, only by the second, by both or by neither. Those found by both or neither
are ignored, and the tool-specific numbers are tested for marginal homogeneity
(i.e. whether their differences give rise to similar totals).

@@

We chose the Tons of Inductive Problems (TIP) benchmark for our ground truth
since it satisfies the criteria specified in $\S$\ref{section:prep}: each
benchmark problem has standalone type and function definitions, making their
separation trivial; known examples from the software verification and inductive
theorem proving literature are included, ensuring relevance to those fields; the
format includes the higher-order functions and inductive datatypes we are
interested in; it is large enough to pose a challenge to current MTE tools; plus
it is accompanied by tooling to convert its custom format (an extension of
SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of languages, including Haskell and
Isabelle.

@@

TIP comes with software to translate its definitions into Haskell and Isabelle
code, including comparison functions and random data generators suitable for
QuickCheck. We translate all 269 unique definitions into a single module/theory
which is imported on each run of the tools, although only those functions which
appear in the current sample are included in the signature and explored. We also
encode all names in hexadecimal to avoid problems with language-specific naming
rules, for example \texttt{add} becomes \texttt{global616464} (the prefix
distinguishes these from local variables and prevents names from beginning with
a digit). This ensures that the generated conjectures will be using the same
names as the ground truth, rather than some lossy variant.

@@

We benchmarked QuickSpec version 0.9.6, a tool written in Haskell for
conjecturing equations involving Haskell functions, described in more detail in
$\S$\ref{sec:existing-tools}. In order to thoroughly benchmark QuickSpec, we
need to automate some of the decisions which are normally left up to the user:

@@

We compare the recall proportions by applying McNemar's test to only those
samples where both tools succeeded. We pooled these together from all sample
sizes to produce table~\ref{table:recall}, and found that the recall of
QuickSpec ($37\%$) is significantly higher than IsaCoSy ($25\%$) with a p-value
of 0.0026.

\begin{table}
  \centering
  \begin{tabular}{ |r|l|l| }
    \hline
    & \bfseries Found by IsaCoSy & \bfseries Not found by IsaCoSy \\
    \hline
    \bfseries Found by QuickSpec     & 28 & 19 \\
    \hline
    \bfseries Not found by QuickSpec & 4  & 75 \\
    \hline
  \end{tabular}
  \caption{Contingency table for ground truth properties, pooled from those
    samples where both QuickSpec and IsaCoSy finished successfully. The combined
    recall for QuickSpec is $37\%$ and for IsaCoSy is $25\%$.}
  \label{table:recall}
\end{table}

@@

\begin{table}
  \centering
  \begin{tabular}{ |r|l|l| }
    \hline
              & \bfseries Interesting & \bfseries Uninteresting \\
    \hline
    \bfseries IsaCoSy   & 32          & 137      \\
    \hline
    \bfseries QuickSpec & 47          & 301      \\
    \hline
  \end{tabular}
  \caption{Interesting and uninteresting conjectures generated by QuickSpec and
    IsaCoSy, pooled from those samples where both both finished successfully.
    The combined precision for QuickSpec is $14\%$ and for IsaCoSy is $19\%$.}
  \label{table:precision}
\end{table}

@@

One notable weakness is that our methodology does not allow negative examples,
such as particularly dull properties that tools should never produce. Everything
outside the ground truth set is treated equally, whether it's genuinely
uninteresting or was only left out due to oversight. In particular this limits
the use of our approach to domains where existing human knowledge surpasses that
discovered by the machine. Any \emph{truly novel} insights discovered during
testing will not, by definition, appear in any existing corpus, and we would in
fact \emph{penalise} tools for such output. We do not believe this to be a
realistic problem for the time being, as long as evaluation is limited to
well-studied domains and results can be generalised to real areas of
application. This does emphasise the continued importance of testing these tools
with real human users, rather than relying solely on artificial benchmarks.
